Magnetized accretion disks around Kerr black holes with scalar hair - I. Constant angular momentum

### 2018-06-28 paper_prd.tex
Some more work on the paper.

\author{Sergio Gimeno-Soler}
\affiliation{Departamento de
  Astronom\'{\i}a y Astrof\'{\i}sica, Universitat de Val\`encia,
  Dr. Moliner 50, 46100, Burjassot (Val\`encia), Spain}

\author{Jos\'e A. Font}
\affiliation{Departamento de
  Astronom\'{\i}a y Astrof\'{\i}sica, Universitat de Val\`encia,
  Dr. Moliner 50, 46100, Burjassot (Val\`encia), Spain}
\affiliation{Observatori Astron\`omic, Universitat de Val\`encia, C/ Catedr\'atico 
  Jos\'e Beltr\'an 2, 46980, Paterna (Val\`encia), Spain}

@@

The organization of this paper is as follows:

@@

\section{Framework}

We use the stationary and axisymmetric metric ansatz provided by\cite{Herdeiro:2014a}

\begin{equation}
\begin{split}
\mathrm{d}s^2 = e^{2F_1}\left(\frac{\mathrm{d}r^2}{N} + r^2\mathrm{d}\theta^2\right) + \\ e^{2F_2}r^2\sin^2 \theta(\mathrm{d}\phi-W\mathrm{d}t)^2 -  e^{2F_0}N\mathrm{d}t^2,
\end{split}
\end{equation}

with $N = 1 - r_H/r$, where $r_H$ is the radius of the event horizon of the BH and $W$, $F_1$, $F_2$, $F_0$ are functions of $r$ and $\theta$.

This metric is a solution to the Einstein-Klein-Gordon field equations $R_{ab} - \frac{1}{2}R g _{ab} = 8 \pi G (T_{\mathrm{SF}})_{ab}$ with

\begin{equation}\label{eq:e-m_scalaf_field}
\begin{split}
(T_{\mathrm{SF}})_{ab} = \partial_a \Psi^* \partial_b \Psi + \partial_b \Psi^* \partial_a \Psi \\ - g_{ab} \left(\frac{1}{2} g^{cd}(\partial_c \Psi^* \partial_d \Psi + \partial_d \Psi^* \partial_c \Psi) + \mu^2 \Psi^* \Psi \right),
\end{split}
\end{equation}

where $\Psi$ is a complex Klein-Gordon field such as $\Psi = \phi(r, \theta) e^{\mathrm{i}m\phi - w t}$ where $w$ is the scalar field frequency and $m$ is the azimuthal harmonic index. Details about the obtention of the solution and the behaviour of the metric functions ${W, F_i}$ are presented in~\cite{Herdeiro:2015b}

@@

\subsection{Distribution of angular momentum and equations of motion}
We introduce the specific angular momentum $l$ and the angular velocity $\Omega$ employing the standard definitions,
\begin{equation}
l = - \frac{u_{\phi}}{u_t}, \;\;\; \Omega = \frac{u^{\phi}}{u^t},
\end{equation}
where $u^{\mu}$ is the fluid four-velocity.
The relationship between $l$ and $\Omega$ is given by the equations
\begin{equation}
l = - \frac{\Omega g_{\phi\phi} + g_{t\phi}}{\Omega g_{t\phi} + g_{tt}}, \;\;\; \Omega = - \frac{l g_{tt} + g_{t\phi}}{l g_{t\phi} + g_{\phi\phi}},
\end{equation}
where we are assuming circular motion, i.e. the four-velocity can be written as
\begin{equation}
u^{\mu} = (u^t, 0, 0, u^{\phi})\,.
\end{equation}

@@

\subsection{Magnetized disks}
We use the procedure described by~\cite{Montero:2007}, where we write the equations of ideal general relativistic MHD as the following conservation laws, $\nabla_{\mu} T^{\mu\nu} = 0$, $\nabla_{\mu} \,^\ast F^{\mu\nu} = 0$, and 
$\nabla_{\mu} (\rho u^{\mu}) = 0$, 
where $\nabla_{\mu}$ is the covariant derivative and
\begin{equation}\label{eq:e-m_tensor}
T^{\mu\nu} = (\rho h + b^2)u^{\mu}u^{\nu} + \left(p + \frac{1}{2}b^2\right)g^{\mu\nu} - b^{\mu}b^{\nu},
\end{equation}
is the energy-momentum tensor of a magnetised perfect fluid, $h$, $\rho$ $p$ being the fluid specific enthalpy, density and fluid pressure, respectively. 
Moreover, $^\ast F^{\mu\nu} = b^{\mu}u^{\nu} - b^{\nu}u^{\mu}$ is the (dual of the) Faraday tensor relative to an observer with 
four-velocity $u^{\mu}$, and $b^{\mu}$ is the magnetic field in that frame, with
$b^2=b^{\mu}b_{\mu}$. Assuming the magnetic field is purely azimuthal, i.e.~$b^r = b^{\theta} = 0$,
and taking into account that the flow is stationary and axisymmetric, the conservation of the current density and of the Faraday tensor follow. Contracting the divergence of Eq.~\eqref{eq:e-m_tensor} with the projection tensor $h^{\alpha}_{\,\,\beta} = \delta^{\alpha}_{\,\,\beta} + u^{\alpha}u_{\beta}$, we arrive at
\begin{equation}
(\rho h + b^2)u_{\nu}\partial_i u^{\nu} + \partial_i\left(p + \frac{b^2}{2}\right) - b_{\nu}\partial_i b^{\nu}=0\,,
\end{equation}
where $i = r, \theta$. Then, we rewrite this equation in terms of the specific angular momentum $l$ and of the angular velocity $\Omega$, to obtain
\begin{equation}\label{eq:diff_ver}
\partial_i(\ln u_t|) - \frac{\Omega \partial_i l}{1-l\Omega} + \frac{\partial_i p}{w} + \frac{\partial_i(\mathcal{L}b^2)}{2\mathcal{L}w} = 0\,,
\end{equation}
where $\mathcal{L} = g_{t\phi}^2 - g_{tt}g_{\phi\phi}$.

@@

%--------------------------------------------------------------------
\section{Method}

\subsection{Building the disk}

@@

\subsection{Numerical method}
We start from a uniform numerical grid $(X, \theta)$ (with $x$ being a compactified radial coordinate such as $X = x/(1+x)$ and $x = \sqrt{r^2 - r_{\mathrm{H}}^2}$) with a domain $[0, 1] \times [0, \pi/2]$ and a number of points of $N_X \times N_\theta = 251 \times 30$. Then, we transform and interpolate the grid to end with a non-uniform grid $(r, \theta)$ with a domain $[r_{\mathrm{H}}, 199.2] \times [0, pi/2]$ and a number of points $N_r \times N_\theta = 2500 \times 300$, this is the grid we use for our computations.

%--------------------------------------------------------------------
\section{Results}
In table~\ref{models_list} we show the different KBHsSH models we will use. As it can be seen, the models go from a Kerr-like model (almost all the mass and angular momentum are stored in BH) to a KBHSH with almost all the mass and angular momentum stored in the scalar field. It is also worth mentioning that, as shown in table~\ref{models_spin_vel} some of the models violate the Kerr bound (in terms of ADM or horizon quantities). This is not worrying because, as shown in~\cite{Herdeiro:2015c} the horizon linear velocity $v_{\mathrm{H}}$ never exceed $1$. For comparison, we also show the spin parameter $a_{\mathrm{H_{eq}}}$ corresponding to a Kerr BH with horizon linear velocity $v_{\mathrm{H}}$.

@@

\begin{table}
\caption{List of models of KBHsSH.}        
\label{models_list}      
\centering          
\begin{tabular}{c c c c  c c c c}
\hline\hline       
 Model & $M_{\mathrm{ADM}}$ & $J_{\mathrm{ADM}}$ & $M_{\mathrm{H}}$ &  $J_{\mathrm{H}}$ & $M_{\mathrm{SF}}$ & $J_{\mathrm{SF}}$ & $r_{\mathrm{H}}$ \\ 
\hline           
I & $0.415$ & $0.172$ & $0.393$ &  $0.15$  & $0.022$ & $0.022$ & $0.2$\\ 
 \hline 
II & $0.630$ & $0.403$ & $0.340$ &  $0.121$  & $0.063$ & $0.282$ & $0.221$ \\
 \hline 
III & $0.797$ & $0.573$ & $0.365$ &  $0.172$  & $0.573$ & $0.432$ & $0.111$ \\ 
 \hline 
IV & $0.933$ & $0.739$ & $0.234$ &  $0.114$  & $0.699$ & $0.625$ & $0.1$ \\ 
 \hline 
V & $0.940$ & $0.757$ & $0.159$ &  $0.076$  & $0.757$ & $0.781$ & $0.091$ \\ 
 \hline 
VI & $0.959$ & $0.795$ & $0.087$ &  $0.034$  & $0.872$ & $0.781$ & $0.088$ \\ 
 \hline 
VII & $0.975$ & $0.85$ & $0.018$ &  $0.002$  & $0.957$ & $0.848$ & $0.04$ \\ 
\hline      
\end{tabular}
\end{table}

@@

%--------------------------------------------------------------------
\section{Conclusions}
Future work: Non-constant angular momentum case, Proca hair, shadows of the system HBH+disk.

>>>>

\author{Sergio Gimeno-Soler}
\affiliation{Departamento de
  Astronom\'{\i}a y Astrof\'{\i}sica, Universitat de Val\`encia,
  C/ Dr. Moliner 50, 46100, Burjassot (Val\`encia), Spain}

\author{Jos\'e A. Font}
\affiliation{Departamento de
  Astronom\'{\i}a y Astrof\'{\i}sica, Universitat de Val\`encia,
  C/ Dr. Moliner 50, 46100, Burjassot (Val\`encia), Spain}
\affiliation{Observatori Astron\`omic, Universitat de Val\`encia, C/ Catedr\'atico 
  Jos\'e Beltr\'an 2, 46980, Paterna (Val\`encia), Spain}

@@

The organization of this paper is as follows: Section~\ref{framework} presents  the  analytic  mathematical framework to build non-self-gravitating magnetized discs in the spacetime of KBHsSH. Section~\ref{procedure} discusses the corresponding numerical methodology. The sequences of equilibrium models are presented and  discussed in Section~\ref{results}. Finally, our conclusions  are  summarized  in  Section~\ref{conclusions}.

@@

%%%%%%%%%%%
\section{Framework}
\label{framework}
%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spacetime metric and KBHsSH models}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

The KBHsSH models we use in this study are built following the procedure described in~\cite{Herdeiro:2015b}. The underlying theoretical framework is the Einstein-Klein-Gordon (EKG) field theory, describing a massive complex scalar field $\Psi$ minimally coupled to Einstein gravity. KBHsSH solutions are obtained by using the following ansatz for the metric and the scalar field~\cite{Herdeiro:2014a}
\begin{eqnarray}
%\begin{split}
\mathrm{d}s^2 &=& e^{2F_1}\left(\frac{\mathrm{d}r^2}{N} + r^2\mathrm{d}\theta^2\right) +  e^{2F_2}r^2\sin^2 \theta(\mathrm{d}\phi-W\mathrm{d}t)^2 
\nonumber \\ 
&-&  e^{2F_0}N\mathrm{d}t^2\,,
\\
%\end{eqnarray}
%\begin{eqnarray}
\Psi &=& \phi(r, \theta) e^{\mathrm{i}(m\varphi - \omega t)} \,,
%\end{split}
\end{eqnarray}
with $N = 1 - r_H/r$, where $r_H$ is the radius of the event horizon of the BH, and $W$, $F_1$, $F_2$, $F_0$ are functions of $r$ and $\theta$. Moreover, $\omega$ is the scalar field frequency and $m$ is the azimuthal harmonic index.

The stationary and axisymmetric metric ansatz is a solution to the EKG field equations $R_{ab} - \frac{1}{2}R g _{ab} = 8 \pi G (T_{\mathrm{SF}})_{ab}$ with 
\begin{eqnarray}\label{eq:e-m_scalaf_field}
(T_{\mathrm{SF}})_{ab} &=& \partial_a \Psi^* \partial_b \Psi + \partial_b \Psi^* \partial_a \Psi 
\nonumber \\ 
&-& g_{ab} \left(\frac{1}{2} g^{cd}(\partial_c \Psi^* \partial_d \Psi + \partial_d \Psi^* \partial_c \Psi) + \mu^2 \Psi^* \Psi \right),
\end{eqnarray}
where $\mu$ is the mass of the scalar field and $^*$ denotes complex conjugation. The interested reader is addressed to~\cite{Herdeiro:2015b} for details on the equations of motion for the scalar field $\Psi$ and the four metric functions ${W, F_i}$, along with their solution.

Table~\ref{models_list} lists the seven KBHsSH models we use in this work. As it can be seen, the models go from a Kerr-like model (almost all the mass and angular momentum are stored in BH) to a KBHSH with almost all the mass and angular momentum stored in the scalar field. It is also worth mentioning that, as shown in table~\ref{models_spin_vel} some of the models violate the Kerr bound (in terms of ADM or horizon quantities). This is not worrying because, as shown in~\cite{Herdeiro:2015c} the horizon linear velocity $v_{\mathrm{H}}$ never exceed $1$. For comparison, we also show the spin parameter $a_{\mathrm{H_{eq}}}$ corresponding to a Kerr BH with horizon linear velocity $v_{\mathrm{H}}$.

%%%%%%%%%%%%%%%%%%%
\begin{table}
\caption{List of models of KBHsSH.}        
\label{models_list}      
\centering          
\begin{tabular}{c c c c  c c c c}
\hline\hline       
 Model & $M_{\mathrm{ADM}}$ & $J_{\mathrm{ADM}}$ & $M_{\mathrm{H}}$ &  $J_{\mathrm{H}}$ & $M_{\mathrm{SF}}$ & $J_{\mathrm{SF}}$ & $r_{\mathrm{H}}$ \\ 
\hline           
I & $0.415$ & $0.172$ & $0.393$ &  $0.15$  & $0.022$ & $0.022$ & $0.2$\\ 
 \hline 
II & $0.630$ & $0.403$ & $0.340$ &  $0.121$  & $0.063$ & $0.282$ & $0.221$ \\
 \hline 
III & $0.797$ & $0.573$ & $0.365$ &  $0.172$  & $0.573$ & $0.432$ & $0.111$ \\ 
 \hline 
IV & $0.933$ & $0.739$ & $0.234$ &  $0.114$  & $0.699$ & $0.625$ & $0.1$ \\ 
 \hline 
V & $0.940$ & $0.757$ & $0.159$ &  $0.076$  & $0.757$ & $0.781$ & $0.091$ \\ 
 \hline 
VI & $0.959$ & $0.795$ & $0.087$ &  $0.034$  & $0.872$ & $0.781$ & $0.088$ \\ 
 \hline 
VII & $0.975$ & $0.85$ & $0.018$ &  $0.002$  & $0.957$ & $0.848$ & $0.04$ \\ 
\hline      
\end{tabular}
\end{table}
%%%%%%%%%%%%%%%%%%%

\begin{figure}
\centering
\includegraphics[scale=0.7]{figures/existence-eps-converted-to.pdf}
\caption{The figure has to be redone to label the 7 models used in this work.}
\label{existence}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Distribution of angular momentum and equations of motion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We introduce the specific angular momentum $l$ and the angular velocity $\Omega$ employing the standard definitions,
\begin{equation}
l = - \frac{u_{\phi}}{u_t}, \;\;\; \Omega = \frac{u^{\phi}}{u^t},
\end{equation}
where $u^{\mu}$ is the fluid four-velocity.
The relationship between $l$ and $\Omega$ is given by the equations
\begin{equation}
l = - \frac{\Omega g_{\phi\phi} + g_{t\phi}}{\Omega g_{t\phi} + g_{tt}}, \;\;\; \Omega = - \frac{l g_{tt} + g_{t\phi}}{l g_{t\phi} + g_{\phi\phi}},
\end{equation}
where we are assuming circular motion, i.e. the four-velocity can be written as
\begin{equation}
u^{\mu} = (u^t, 0, 0, u^{\phi})\,.
\end{equation}

@@

%%%%%%%%%%%%%%%%
\subsection{Magnetized disks}
%%%%%%%%%%%%%%%%

We use the procedure described by~\cite{Montero:2007}, where we write the equations of ideal general relativistic MHD as the following conservation laws, $\nabla_{\mu} T^{\mu\nu} = 0$, $\nabla_{\mu} \,^\ast F^{\mu\nu} = 0$, and 
$\nabla_{\mu} (\rho u^{\mu}) = 0$, 
where $\nabla_{\mu}$ is the covariant derivative and
\begin{equation}\label{eq:e-m_tensor}
T^{\mu\nu} = (\rho h + b^2)u^{\mu}u^{\nu} + \left(p + \frac{1}{2}b^2\right)g^{\mu\nu} - b^{\mu}b^{\nu},
\end{equation}
is the energy-momentum tensor of a magnetised perfect fluid, $h$, $\rho$ $p$ being the fluid specific enthalpy, density and fluid pressure, respectively. 
Moreover, $^\ast F^{\mu\nu} = b^{\mu}u^{\nu} - b^{\nu}u^{\mu}$ is the (dual of the) Faraday tensor relative to an observer with 
four-velocity $u^{\mu}$, and $b^{\mu}$ is the magnetic field in that frame, with
$b^2=b^{\mu}b_{\mu}$. Assuming the magnetic field is purely azimuthal, i.e.~$b^r = b^{\theta} = 0$,
and taking into account that the flow is stationary and axisymmetric, the conservation of the current density and of the Faraday tensor follow. Contracting the divergence of Eq.~\eqref{eq:e-m_tensor} with the projection tensor $h^{\alpha}_{\,\,\beta} = \delta^{\alpha}_{\,\,\beta} + u^{\alpha}u_{\beta}$, we arrive at
\begin{equation}
(\rho h + b^2)u_{\nu}\partial_i u^{\nu} + \partial_i\left(p + \frac{b^2}{2}\right) - b_{\nu}\partial_i b^{\nu}=0\,,
\end{equation}
where $i = r, \theta$. Then, we rewrite this equation in terms of the specific angular momentum $l$ and of the angular velocity $\Omega$, to obtain
\begin{equation}\label{eq:diff_ver}
\partial_i(\ln u_t|) - \frac{\Omega \partial_i l}{1-l\Omega} + \frac{\partial_i p}{w} + \frac{\partial_i(\mathcal{L}b^2)}{2\mathcal{L}w} = 0\,,
\end{equation}
where $\mathcal{L} = g_{t\phi}^2 - g_{tt}g_{\phi\phi}$.

@@

%%%%%%%%%%%%
\section{Methodology}
\label{procedure}
%%%%%%%%%%%%

%%%%%%%%%%%%%%%%
\subsection{Building the disk}
%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%
\subsection{Numerical method}
%%%%%%%%%%%%%%%%%

We start from a uniform numerical grid $(X, \theta)$ (with $x$ being a compactified radial coordinate such as $X = x/(1+x)$ and $x = \sqrt{r^2 - r_{\mathrm{H}}^2}$) with a domain $[0, 1] \times [0, \pi/2]$ and a number of points of $N_X \times N_\theta = 251 \times 30$. Then, we transform and interpolate the grid to end with a non-uniform grid $(r, \theta)$ with a domain $[r_{\mathrm{H}}, 199.2] \times [0, pi/2]$ and a number of points $N_r \times N_\theta = 2500 \times 300$, this is the grid we use for our computations.

%%%%%%%%%%
\section{Results}
\label{results}
%%%%%%%%%%

%In table~\ref{models_list} we show the different KBHsSH models we will use. As it can be seen, the models go from a Kerr-like model (almost all the mass and angular momentum are stored in BH) to a KBHSH with almost all the mass and angular momentum stored in the scalar field. It is also worth mentioning that, as shown in table~\ref{models_spin_vel} some of the models violate the Kerr bound (in terms of ADM or horizon quantities). This is not worrying because, as shown in~\cite{Herdeiro:2015c} the horizon linear velocity $v_{\mathrm{H}}$ never exceed $1$. For comparison, we also show the spin parameter $a_{\mathrm{H_{eq}}}$ corresponding to a Kerr BH with horizon linear velocity $v_{\mathrm{H}}$.

@@

%\begin{table}
%\caption{List of models of KBHsSH.}        
%\label{models_list}      
%\centering          
%\begin{tabular}{c c c c  c c c c}
%\hline\hline       
% Model & $M_{\mathrm{ADM}}$ & $J_{\mathrm{ADM}}$ & $M_{\mathrm{H}}$ &  $J_{\mathrm{H}}$ & $M_{\mathrm{SF}}$ & $J_{\mathrm{SF}}$ & $r_{\mathrm{H}}$ \\ 
%\hline           
%I & $0.415$ & $0.172$ & $0.393$ &  $0.15$  & $0.022$ & $0.022$ & $0.2$\\ 
% \hline 
%II & $0.630$ & $0.403$ & $0.340$ &  $0.121$  & $0.063$ & $0.282$ & $0.221$ \\
% \hline 
%III & $0.797$ & $0.573$ & $0.365$ &  $0.172$  & $0.573$ & $0.432$ & $0.111$ \\ 
% \hline 
%IV & $0.933$ & $0.739$ & $0.234$ &  $0.114$  & $0.699$ & $0.625$ & $0.1$ \\ 
% \hline 
%V & $0.940$ & $0.757$ & $0.159$ &  $0.076$  & $0.757$ & $0.781$ & $0.091$ \\ 
% \hline 
%VI & $0.959$ & $0.795$ & $0.087$ &  $0.034$  & $0.872$ & $0.781$ & $0.088$ \\ 
% \hline 
%VII & $0.975$ & $0.85$ & $0.018$ &  $0.002$  & $0.957$ & $0.848$ & $0.04$ \\ 
%\hline      
%\end{tabular}
%\end{table}
%
%\begin{table}
%\caption{Values of the normalized spin parameter for the ADM quantities ($a_{\mathrm{ADM}}$), for the BH horizon quantities ($a_{\mathrm{H}}$), the horizon linear velocity ($v_{\mathrm{H}}$) and the spin parameter corresponding to a KBH with a linear velocity equal to $v_{\mathrm{H}}$, ($a_{\mathrm{H_{eq}}}$).}        
%\label{models_spin_vel}      
%\centering          
%\begin{tabular}{c c c c c}
%\hline\hline       
% Model & $a_{\mathrm{ADM}}$ & $a_{\mathrm{H}}$ & $v_{\mathrm{H}}$ & $a_{\mathrm{H_{eq}}}$ \\ 
%\hline           
%I & $0.9987$ & $0.9712$ & $0.7685$ & $0.9663$ \\ 
% \hline 
%II & $1.014$ & $0.3760$ & $0.6802$ & $0.9301$ \\
% \hline 
%III & $0.9032$ & $1.295$ & $0.7524$ & $0.9608$ \\ 
% \hline 
%IV & $0.8489$ & $2.082$ & $0.5635$ & $0.8554$ \\ 
% \hline 
%V & $0.8560$ & $3.017$ & $0.4438$ & $0.7415$ \\ 
% \hline 
%VI & $0.9477$ & $3.947$ & $0.2988$ & $0.5487$ \\ 
% \hline 
%VII & $0.8941$ & $6.173$ & $0.09732$ & $0.1928$ \\ 
%\hline      
%\end{tabular}
%\end{table}

@@

%%%%%%%%%%%%
\section{Conclusions}
\label{conclusions}
%%%%%%%%%%%%

Future work: Non-constant angular momentum case, Proca hair, shadows of the system HBH+disk.
